Mechanising Hylomorphisms for Extraction of Executable Code

### 2025-03-18 paper/paper.tex
minor

\paragraph*{Recursive Anamorphisms}
Anamorphisms applied to recursive coalgebras specialise to hylomorphisms into an
inductive data type in the following way. A recursive coalgebra can be applied
only finitely many times, therefore when this is applied to an anamorphism the
only possible types it can produce from the seed are the finite ones. We call
this special kind of recursion scheme \emph{recursive anamorphism}. We can show
that recursive anamorphisms of type $X \to \nu F$ can also be given the type
$X \to \mu F$. Moreover, these anamorphisms are exactly hylomorphisms on the
recursive $F$-coalgebra and the algebra $\InIso$ for the inductive data type
$\mu F$. This fact falls out from the uniqueness property of the hylomorphism
and the fact that recursive anamorphisms satisfy the same equation.

@@

Note however, that the type $\forall s : S.\; P\;s \to X$ would force us to add
the functional extensionality axiom to deal with the heterogeneous equality of
positions. Thus, we require that all the types used in our definitions are
\emph{setoids}, where we will use \coq{x =e y} to represent setoid equality. All
morphisms need to be \emph{proper} morphisms with respect to the function
\emph{respectfulness} relation. In our development, we use
\begin{coqcode}
Notation "A ~> B" := (* proper morphisms  s.t. forall x y, x =e y -> f x =e f y *).
\end{coqcode}
We add an automatic coercion from \coq{A ~> B} to \coq{A -> B} and provide
tactics to automatically discharge proofs that morphisms are proper,  for some
simple cases.

>>>>

\paragraph{Recursive Anamorphisms}
Anamorphisms applied to recursive coalgebras specialise to hylomorphisms into an
inductive data type in the following way. A recursive coalgebra can be applied
only finitely many times, therefore when this is applied to an anamorphism the
only possible types it can produce from the seed are the finite ones. We call
this special kind of recursion scheme \emph{recursive anamorphism}. We can show
that recursive anamorphisms of type $X \to \nu F$ can also be given the type
$X \to \mu F$. Moreover, these anamorphisms are exactly hylomorphisms on the
recursive $F$-coalgebra and the algebra $\InIso$ for the inductive data type
$\mu F$. This fact falls out from the uniqueness property of the hylomorphism
and the fact that recursive anamorphisms satisfy the same equation.

@@

Note however, that the type $\forall s : S.\; P\;s \to X$ would force us to add
the functional extensionality axiom to deal with the heterogeneous equality of
position. Thus, we require that all the types used in our definitions are
\emph{setoids}, where we will use \coq{x =e y} to represent setoid equality. All
morphisms need to be \emph{proper} morphisms with respect to the function
\emph{respectfulness} relation. In our development, we use
\begin{coqcode}
Notation "A ~> B" := (* proper morphisms  s.t. forall x y, x =e y -> f x =e f y *).
\end{coqcode}
We add an automatic coercion from \coq{A ~> B} to \coq{A -> B} and provide tactics
to automatically discharge proofs that \coq{forall x y, x =e y -> f x =e f y}
for the simplest cases of \coq{f}.
